---
title: "Assignment 2 - Data 621"
author: "Coco Donovan"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading Packages:

```{r}
library(readr)
library(knitr)
library(ggplot2)
library(pROC)
library(caret)
library(glue)
```

### 1. 

Download the classification output data set (attached in Blackboard to the 
assignment).

```{r}
df <- read_csv('classification-output-data.csv')
```

### 2.

The data set has three key columns we will use:
  * **class**: the actual class for the observation
  * **scored.class**: the predicted class for the observation (based on a threshold of
  0.5)
  * **scored.probability**: the predicted probability of success for the observation
  
  
Use the table() function to get the raw confusion matrix for this scored dataset.
Make sure you understand the output. In particular, do the rows represent the actual
or predicted class? The columns?


**Answer:**


Confusion Matrices generally follow this format:

```{r}
# Create the confusion matrix data
confusion_matrix <- matrix(c("TN", "FN", "FP", "TP"), nrow = 2, byrow = TRUE)
rownames(confusion_matrix) <- c("Predicted Negative", "Predicted Positive")
colnames(confusion_matrix) <- c("Actual Negative", "Actual Positive")

# Print the confusion matrix using kable
kable(confusion_matrix, format = "markdown")
```

This would be our raw confusion matrix:

```{r}
confusion_matrix <- table(df$scored.class, df$class)
confusion_matrix_with_totals <- addmargins(confusion_matrix)
print(kable(confusion_matrix_with_totals, format = "markdown"))
```

Note: Here 0 is just the binary form of "Negative" and 1 is just the binary form
"Positive."


Based on the general format and my raw confusion matrix, I can say that 119 cases
that were actually negative were predicted as negative, whereas 5 of those cases
that were actually negative were predicted as positive. For the cases that were 
actually positive, 30 were incorrectly predicted as negative, whereas the other 
27 cases associated with an actual positive were correctly predicted as positive.


### 3.

Write a function that takes the data set as a dataframe, with actual and 
predicted classifications identified, and returns the accuracy of the predictions.

Accuracy = $\frac{TP+TN}{TP+FP+TN+FN}$

```{r}
Accuracy = function(df) {
  
  TP = sum(df$class == 1 & df$scored.class == 1)
  TN = sum(df$class == 0 & df$scored.class == 0)
  FP = sum(df$class == 0 & df$scored.class == 1)
  FN = sum(df$class == 1 & df$scored.class == 0)
  
  return((TP+TN)/(TP+FP+TN+FN))
}
```

### 4.

Write a function that takes the data set as a dataframe, with actual and predicted
classifications identified, and returns the classification error rate of the predictions.

Classification Error Rate = $\frac{FP+FN}{TP+FP+TN+FN}$

```{r}
Classification_Error_Rate = function(df) {
  
  TP = sum(df$class == 1 & df$scored.class == 1)
  TN = sum(df$class == 0 & df$scored.class == 0)
  FP = sum(df$class == 0 & df$scored.class == 1)
  FN = sum(df$class == 1 & df$scored.class == 0)
  
  return((FP+FN)/(TP+FP+TN+FN))
}
```

Verify that you get an accuracy and an error rate that sums to one.


**Answer:**


Classification Error Rate is represented by the expression $\frac{FP+FN}{TP+FP+TN+FN}$
and accuracy is represented by the expression $\frac{TP+TN}{TP+FP+TN+FN}$. To check
if the two sum to 1, I can just add the expressions, like so:

$(\frac{FP+FN}{TP+FP+TN+FN}) + (\frac{TP+TN}{TP+FP+TN+FN})$

This gives me the new fraction:

$\frac{TP+FP+TN+FN}{TP+FP+TN+FN}$

This fraction has the same numerator and denominator, meaning that the sum of Accuracy
and Classification Error Rate should sum to 1, and if that is not the case, then
something is wrong with the derived calculations.

Later in this analysis, I will check to see if my calculated Classification Error
Rate and Accuracy sum to 1.

### 5.
Write a function that takes the data set as a dataframe, with actual and predicted
classifications identified, and returns the precision of the predictions.

Precision = $\frac{TP}{TP+FP}$

```{r}
Precision = function(df) {
  
  TP = sum(df$class == 1 & df$scored.class == 1)
  FP = sum(df$class == 0 & df$scored.class == 1)
  
  return((TP)/(TP+FP))
}
```

### 6.
Write a function that takes the data set as a dataframe, with actual and predicted
classifications identified, and returns the sensitivity of the predictions. 
Sensitivity is also known as recall.

Sensitivity = $\frac{TP}{TP+FN}$

```{r}
Sensitivity = function(df) {
  
  FN <- sum(df$class == 1 & df$scored.class == 0)
  TP <- sum(df$class == 1 & df$scored.class ==1)
  
  return((TP)/(TP+FN))
}

Sensitivity(df)
```

### 7. 
Write a function that takes the data set as a dataframe, with actual and predicted
classifications identified, and returns the specificity of the predictions.

Specificity = $\frac{TN}{TN+FP}$

```{r}
Specificity = function(df) {
  
  TP = sum(df$class == 1 & df$scored.class == 1)
  TN = sum(df$class == 0 & df$scored.class == 0)
  FP = sum(df$class == 0 & df$scored.class == 1)
  FN = sum(df$class == 1 & df$scored.class == 0)
  
  return((TN)/(TN+FP))
}
```

### 8. 
Write a function that takes the data set as a dataframe, with actual and predicted
classifications identified, and returns the F1 score of the predictions.

F1 Score = $\frac{2 \times Precision \times Sensitivity}{Precision+Sensitivity}$

```{r}
F1 = function (df) {
  f1 = (2 * Precision(df) * Sensitivity(df)) / (Precision(df) + Sensitivity(df))
  return(f1)
}
```

### 9.
Before we move on, let’s consider a question that was asked: What are the bounds
on the $F1$ score? Show that the $F1$ score will always be between 0 and 1. 
(Hint: If $0 < a < 1$ and $0 < b < 1$ then $ab < a$.)


**Answer**


Sensitivity = $\frac{TP}{TP+FN}$ and Precision = $\frac{TP}{TP+FP}$, where 
$TP, FP and FN$ are non-negative values. From that fact that all the values in these
formulas are non-negative we can tell that:


$0 \leq Sensitivity \leq 1$ 


and 


$0 \leq Precision \leq 1$.


It is true that if you multiply a number, $a$, by another number, $b$ that is in
between 0 and 1, the product of the two numbers will be at most a or b (depending
on which number is greater). To show that this is true, I'll show what happens if
b is 0 (the minimum value it can take on) and if b is 1 (the maximum value it can
take on).


When $b=0$:


$a\times b$ = $a\times 0$ = 0


When $b=1$:


$a\times b$ = $a\times 1$ = a


So, we can say that $ab\leq a$ and using this logic (given that a operates under
the same between 0 and 1 condition) we can also say that $ab\leq b$. Further, 
I can then say that $2\times ab \leq a+b$ (coming as a direct result of combining
both of these inequalities).


Now, Precision and Sensitivity operate under the same between 0 and 1 condition
as both $a$ and $b$, so I can then say that $Precision\times Sensitivity \leq Sensitivity$
and that $Precision\times Sensitivity \leq Precision$.


Using the combination of inequalities I employed above, I can say that:


$2 \times Precision\times Sensitivity \leq Precision + Sensitivity$


The Left side of this inequality is the numerator of the F1 Score function and the 
right side of the inequality is the denominator of the F1 Score function. What I 
have shown is the the denominator will always be as large or larger than the 
numerator (meaning the maximum value it can take on is 1).


Therefore, because the F1 Score is positive and the maximum value it can take on
is 1, the F1 Score will always be between 0 and 1.

### 10.
Write a function that generates an ROC curve from a data set with a true 
classification column (class in our example) and a probability column (scored.probability
in our example). Your function should return a list that includes the plot of the
ROC curve and a vector that contains the calculated area under the curve (AUC). 
Note that I recommend using a sequence of thresholds ranging from 0 to 1 at 0.01
intervals.

```{r}

```

### 11.
Use your **created R functions** and the provided classification output data set
to produce all of the classification metrics discussed above.

#### Accuracy:

```{r}
df_accuracy = Accuracy(df)

print(df_accuracy)
```

It seems that this data represents a classification with about 80.6% accuracy.

#### Classification Error Rate:

```{r}
df_cle = Classification_Error_Rate(df)

print(df_cle)
```

It seesm that this data represents a classifcation with a Classification Error Rate
of about 19.3%.

#### Verifying that Accuracy + Classification Error Rate = 1:

```{r}
print((df_cle + df_accuracy) == 1)
```

The accuracy and error rate sum to 1 in this example using the functions I created.

#### Precision:

```{r}
df_precision = Precision(df)
print(df_precision)
```

Based on this calculation, this data represents a classification method with about
84.4% precision.

#### Sensitivity:

```{r}
df_sensitivity = Sensitivity(df)
print(df_sensitivity)
```

Based on this calculation, this data represents a classification method with about
47.4% sensitivity.

#### Specificity:

```{r}
df_specificity = Specificity(df)
print(df_specificity)
```

Based on this calculation, this data represents a classification method with about
96% specificity.

#### F1 Score:

```{r}
F1(df)
```

It seems that this data represents a classification with an F1 Score of 0.6067416.

### 12.
Investigate the **caret package**. In particular, consider the functions confusionMatrix,
sensitivity, and specificity. Apply the functions to the data set. How do the results
compare with your own functions?

```{r}
caret = confusionMatrix(as.factor(df$scored.class), as.factor(df$class), positive = "1")

print(glue("caret: {caret$overall[['Accuracy']]}"))
print(glue("My calculation: {Accuracy(df)}"))
```

The value I calculated for Accuracy is the same as the value produced using the 
confusionMatrix package.

```{r}
print('Sensitivity')
print(glue("confusionMatrix: {caret$byClass[['Sensitivity']]}"))
print(glue("My calculation: {Sensitivity(df)}"))
print(glue('sensitivity function: {sensitivity(as.factor(df$scored.class), as.factor(df$class), positive = "1")}'))
```

The value I calculated for Sensitivity is the same as the value produced using the 
confusionMatrix and sensitivity functions.

```{r}
print('Specificity')
print(glue("caret: {caret$byClass[['Specificity']]}"))
print(glue("My calculation: {Specificity(df)}"))
print(glue('specificity function: {specificity(as.factor(df$scored.class), as.factor(df$class), negative = 0)}'))
```

The value I calculated for Specificity is the same as the value produced using the 
confusionMatrix and specificity functions.

### 13.
Investigate the **pROC** package. Use it to generate an ROC curve for the data set.
How do the results compare with your own functions?

```{r}
plot(roc(df$class, df$scored.probability), main="ROC Curve from pROC Package")
```




